---
title: "Station Summary Data"
author: "Amanda Liczner"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE, warning=FALSE}
library(motus)
library(lubridate)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(sf)
library(ggplot2)
library(knitr)
library(mapview)
library(DT)
library(ggrepel)
library(sfheaders)
library(ggridges)
library(plotly)

Sys.setenv(TZ = "UTC") 

```

Overview of the data from Motus Project #1 summary data (downloaded off the web). 

### Data preparation summary
The summary data was downloaded from all Motus Project #1 towers.
The data are downloaded as "pairs" with the current and previous station noted

**Additional columns were added:**

- column for season
  - spring = March 1 - June 15
  - summer = June 15 - August 18
  - fall = August 15 - November 15
  - Winter = November 15- April 15
- column for sunrise/sunset time
- column for sunrise/sunset local time
- column for movement duration in hours
- a column for whether the detection is likely a flight (< 12 hrs duration)
or an incident (> 12 hr duration)

**cleaning detections that went back in time**

There are detections where the end time is before the start time. These are some sort of error and need to be removed

## Flight information added

Flight specific information was added as additional columns
This includes:

- flight number (sequential flights per individual)
- speed per step in the flight
- turning angle (where it could be calculated)
- diel period (daylight or night) of the flight
- hours from sunrise and sunset (can be used to identify flights close to take-off or landing)
- additional grouping variable added:
  - which subbasin the occurrence took place in within the GLWS

## Individual information added

Information about each individual was added including:

- the tag deployment coordinates
- sex/age of the individual (if the project included this)
- days post migration the detection occurred
- tagging location potential issue flags
  - whether the tagging location was inside or outside the GLWS
  - how far from the boundary and the closest lake shoreline the tagging location was

## Spring phenology

**Number of detections during spring**

``` {r echo = FALSE}

fullData <- read.csv ("StationPairsFiltered.csv")

springData <- fullData %>% 
  filter(
    season == "Spring Migration"
  )
nrow(springData)
```
**Number of species**

``` {r echo = FALSE}
length(unique(springData$species))
```

**Number of tags**
``` {r echo = FALSE}
length(unique(springData$tagDeployID))
```

### Spring Arrival Phenology by Subbasin Plots

``` {r echo = FALSE}
fullData <- read.csv ("StationPairsFiltered.csv")

springData <- fullData %>% 
  filter(
    season == "Spring Migration",
  )
nrow(springData)

springTable <- springData %>%  
  mutate(
    tsEnd_dt = as_datetime(tsEnd_dt, tz = "GMT"),
    # merging georgian bay with lake huron so they are not separate. 
    subbasin = if_else(subbasin == "geo_bay", "lk_huron", subbasin)
  )

species_subbasin_counts <- springTable  %>%
  group_by(species) %>%
  summarise(
    num_subbasins = n_distinct(subbasin, 
                               na.rm = TRUE),
    basins_present = paste(sort(unique(subbasin)), 
                           collapse = ", "),
    .groups = "drop"
  )

```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=16, fig.width=11}
library(ggplot2)
library(ggridges)
library(dplyr)
library(purrr)
library(tidyr) # Needed for complete()

clean_basin_names <- c(
  "lk_erie" = "Lake Erie",
  "lk_huron"= "Lake Huron",
  "lk_ont"  = "Lake Ontario",
  "lk_sup" = "Lake Superior",
  "lk_mich" = "Lake Michigan")

#summarize original raw counts
all_basins_plot_data <- springTable %>%
  mutate(
    month_day = format(tsEnd_dt, "%m-%d"),
    date = as.Date(paste0("2026-", 
                          month_day))
  ) %>%
  filter(!is.na(subbasin), 
         subbasin != "") %>%
  group_by(subbasin, 
           species, 
           date) %>%
  summarise(count = n(),
            .groups = "drop")

# get start and end dates
start_date <- min(all_basins_plot_data$date, na.rm = TRUE)
end_date   <- max(all_basins_plot_data$date, na.rm = TRUE)

#fills in missing times s
completed_data <- all_basins_plot_data %>%
  group_by(subbasin, species) %>%
  complete(date = seq.Date(start_date,
                           end_date, 
                           by = "1 day"), 
           fill = list(count = 0)) %>%
  ungroup()

subbasins_list <- unique(completed_data$subbasin)

# loop plots
walk(subbasins_list, 
     function(basin_code) {
  
  single_basin_data <- completed_data %>% 
    filter(subbasin == basin_code)
  
  if (nrow(single_basin_data) == 0) return(NULL)
  
  proper_name <- clean_basin_names[basin_code]
  if (is.na(proper_name)) proper_name <- toupper(basin_code)
  
  # plotting
  p <- ggplot(single_basin_data, 
              aes(x = date, 
                  y = species, 
                  height = count, 
                  fill = species)) +
    geom_ridgeline(scale = 0.5, 
                   alpha = 0.6, 
                   color = "grey85",) +
        geom_hline(yintercept = unique(single_basin_data$species), 
               color = "grey85",
               size = 0.4) +
    
    scale_x_date(date_labels = "%b %d", 
                 date_breaks = "2 weeks") +
    
    theme_ridges() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 14, hjust = 1, vjust = 0.5)
    ) +
    labs(
      title = paste0("Spring Phenology: ", proper_name),
      x = "Date", 
      y = NULL
    )
  
  print(p)
})
```