---
title: "Station Summary Data"
author: "Amanda Liczner"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(motus)
library(lubridate)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)

Sys.setenv(TZ = "UTC") 

```

Overview of the data from Motus Project #1 summary data (downloaded off the web). 

### Data preparation summary
The summary data was downloaded from all Motus Project #1 towers.
The data are downloaded as "pairs" with the current and previous station noted

The downloaded data included the following headings:
```{r echo=FALSE}
data_active <- read.csv("./StationDownloads/Motus-data-project-1_detections_downloaded-2026-06-03.csv")
#at first just did active towers to prevent crashing
data_inactive <- read.csv ("./StationDownloads/Motus-data-multiple-stations_detections_downloaded-2026-06-09.csv")

#the inactive data had frequency as num, convert so it matches
data_inactive <- data_inactive %>%  
  mutate(tagFrequency = as.character(tagFrequency))

#join active and inactive stations together

data_raw <- bind_rows(data_active, data_inactive) %>%
  select(-tagFrequency) %>%  # some rows have it some do not
  distinct()
head(data_raw)
```
Additional columns were added such as:

- calculated my own movement_duration_hous, traveDurationHours is from the download
- calculated my own distance_km, travelDistanceKMs is from the download
- added in a column for season
  - spring = April 15 - June 15
  - summer = June 15 - August 18
  - fall = August 15 - November 15
  - Winter = November 15- April 15
- sunrise_utc/sunset_utc: the timing of sunrise/sunset in UTC
- sunrise_local/sunset_local: the timing of sunrise/Sunset in Eastern Time
- current_in_GLWS/previous_in_GLWS: if the station (current or previous) is within the Great Lakes Watershed
- distance_km: the distance between the two stations
- valid_distance: 
  - FALSE_LESS = the distance between stations is less than 30 km
  - FALSE_MORE = the distance between stations is more than 965 km 
  - TRUE = the distance between stations is between 30 - 965 km
- flight:
  flight = when the detection between stations is < 12 hrs
  incidence = when the detection between stations is > 12 hrs
- stationDuplicateFlag: flags detection issues (note buffer of 3 seconds was used to account for potential delays in detection and processing time at stations)
  - flag_1 = the start and end time are the same
  - flag_2 multiple towers detected at once, differing start times between rows going back in time
  - flag_3 mismatched run where it ends before its start time. 
  - flag_1_resolved = where flag_1 issues were corrected 
- eventClusterID = unique identifier assigned to a group of overlapping or continuous tracking events for a specific individual, created while tracking flagged errors
- new_flight = to identify the start of a new flight, used while tracking flagged errrors, a new flight is when the movement duration is > 12 hrs
- flight_number: a sequential numbering of flights for that individual
- flight_ID: adds "Flight_" to the flight number (from flight_number)
- step_speed_kmh: the speed in km/h at that flight step
- turning_angle: the turning angle calculated, where possible (=> detections needed)
- diel_period: whether the flight is occuring at night or during daylight
- nearSun: whether the flight occured within one hour of sunrrise/sunset
  - sunrise = the flight occurred within one hour of sunrise
  - sunset = the flight occurred within on hour of sunset
  - none = the flight did not occur within one hour of sunrise or sunset. 




Data cleaning: 
- Detections at the same current and previous station were removed as we are focusing on flights


```{r }
flightInfo_df <- read.csv("StationPairsFiltered.csv")
str(flightInfo_df)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
