---
title: "Station Summary Data"
author: "Amanda Liczner"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE, warning=FALSE}
library(motus)
library(lubridate)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(sf)
library(ggplot2)
library(knitr)
library(mapview)
library(DT)

Sys.setenv(TZ = "UTC") 

```

Overview of the data from Motus Project #1 summary data (downloaded off the web). 

### Data preparation summary
The summary data was downloaded from all Motus Project #1 towers.
The data are downloaded as "pairs" with the current and previous station noted

A preview downloaded data:
```{r echo=FALSE, warning=FALSE}
data_ON <- read.csv("./StationDownloads/Motus-data-region-CA-ON_detections_downloaded-2026-06-23.csv")
data_NY <- read.csv("./StationDownloads/Motus-data-region-US-NY_detections_downloaded-2026-06-23.csv")
data_PA <- read.csv("./StationDownloads/Motus-data-region-US-PA_detections_downloaded-2026-06-23.csv")
data_OH <- read.csv("./StationDownloads/Motus-data-region-US-OH_detections_downloaded-2026-06-23.csv")
data_IN <- read.csv("./StationDownloads/Motus-data-region-US-IN_detections_downloaded-2026-06-23.csv")
data_IL <- read.csv("./StationDownloads/Motus-data-region-US-IL_detections_downloaded-2026-06-23.csv")
data_WI <- read.csv("./StationDownloads/Motus-data-region-US-WI_detections_downloaded-2026-06-23.csv")

# join all the station data 
data_raw <- bind_rows(data_ON,
                      data_NY,
                      data_PA,
                      data_OH,
                      data_IN,
                      data_IL,
                      data_WI) %>%  # some rows have it some do not
  distinct() 

# Format the date columns using lubridate functions
data_cleaned <- data_raw %>%
  mutate(
    tsStart_dt = dmy_hms(tsStart, tz = "UTC"),
    tsEnd_dt   = dmy_hms(tsEnd, tz = "UTC"),
    year_start = year(tsStart_dt),
    year_end   = year(tsEnd_dt)
    )

#================================================================
# Station metadata json file
#================================================================

# Station metadata json file with GPS coordinates
stationsJSON <- jsonlite::read_json("./JSONFiles/Stations.json")

stationsJSON <- jsonlite::fromJSON("./JSONFiles/Stations.json", 
                                   simplifyDataFrame = TRUE)

readr::write_csv(stationsJSON$results, "StationsJSON.csv")

#getting coordinates from stationsJSON to make a lookup table for 
# previous tower coordinates
id_coords_lookup <- stationsJSON$results %>% 
  select(stationID, 
         latitude, 
         longitude) %>% 
  filter(!is.na(latitude) & !is.na(longitude)) %>% 
  # Convert stationID to integer
  mutate(stationID = as.integer(stationID)) %>% 
  # Group by ID and take the first coordinate 
  group_by(stationID) %>% 
  summarise(
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  )

# Add previous station coordinates using the ID columns
summary_coords <- data_cleaned %>% 
  # Convert both ID columns to integers
  mutate(
    previousStationID = as.integer(previousStationID),
    stationID         = as.integer(stationID)
  ) %>% 
  
  # Join the lookup table to the previousStationID column
  left_join(
    id_coords_lookup, 
    by = c("previousStationID" = "stationID")
  ) %>% 
  
  # Rename the new columns coordinates
  rename(
    lat_previous = latitude,
    lon_previous = longitude
  )

head(summary_coords)
```

**Additional columns were added:**

- column for season
  - spring = April 15 - June 15
  - summer = June 15 - August 18
  - fall = August 15 - November 15
  - Winter = November 15- April 15

```{r echo=FALSE}

obs_sum <- read.csv("observationSummary.csv")

table(obs_sum$season)
```

- column for sunrise/sunset time
- column for sunrise/sunset local time
- column for movement duration in hours

**number of detections in raw data**

```{r echo=FALSE}
nrow(obs_sum)
```

## Data cleaning

Number of detections after removing unknown stations and station pairs that were the same
(bird remained at the same station)

```{r echo=FALSE}
station_pairs <- read.csv("observationSummary.csv") #142303 obs

#filter out same station pairs and station unknown

filtered_pairs <- station_pairs %>% 
  filter(
    previousStationID != stationID,
    previousStationName != "Unknown",
    previousStationName != "Unknown station"
  )
nrow(filtered_pairs)
```

Number of detectons with at least one station in the station-pair within the Great Lakes
Watershed (GLWS)
```{r echo=FALSE}
#loading great lakes watershed polygon, contains subbasins for each lake
GLWatershed <- st_read("./greatlakes_subbasins/greatlakes_subbasins.shp")
#checking CRS
st_crs(GLWatershed) #ESPG 6269

#transform both the polygon and station pairs to lambert conformal conic

#need to run current and previous stations separately
current_sf <- st_as_sf(filtered_pairs, 
                       coords = c("lon", "lat"), 
                       crs = 4326) %>%
  st_transform(crs = 3348)

previous_sf <- st_as_sf(filtered_pairs, 
                        coords = c("lon_previous", "lat_previous"), 
                        crs = 4326) %>%
  st_transform(crs = 3348)

GLWS_proj <- st_transform(GLWatershed, crs = 3348)

# check if either of the stations are within the GL watershed, 
filtered_pairs$current_in_GLWS  <- lengths(st_intersects(current_sf, 
                                                         GLWS_proj)) > 0
filtered_pairs$previous_in_GLWS <- lengths(st_intersects(previous_sf, 
                                                         GLWS_proj)) > 0

# calculate distance between points
filtered_pairs$distance_km <- as.numeric(st_distance(current_sf, 
                                                     previous_sf, 
                                                     by_element = TRUE)
) / 1000 #convert to km


#==================================================================
# station pair filtering by criteria
#==================================================================

filtered_df <- filtered_pairs %>%
  filter(
    # Keep if either the current OR the previous station is in GLWS
    (current_in_GLWS | previous_in_GLWS))

nrow(filtered_df) #35865  obs
```

columns added continued: 

- a column for whether the detection is likely a flight (< 12 hrs duration)
or an incident (> 12 hr duration)

```{r echo = FALSE}

filtered_df$flight <- ifelse(filtered_df$movement_duration_hours < 12,
                             "flight",
                             "incidence") 
table(filtered_df$flight)
```

**cleaning detections that went back in time**

There are detections where the end time is before the start time. These are some sort of error and need to be removed

The number of detections after removing the backwards detections:
``` {r echo = FALSE}
cleaned_backwards <- filtered_df %>%
  mutate(
    tsStart_dt = ymd_hms(tsStart_dt, tz = "UTC"),
    tsEnd_dt = ymd_hms(tsEnd_dt, tz = "UTC")
  ) %>%
  filter(tsEnd_dt >= tsStart_dt)
nrow(cleaned_backwards)
```
**cleaning detections that are false positives and local overlaps**

Two other data problems exist:

- false positives
- local overlapping stations picking up the same individual

local overlap was distinguished from false positives by comparing speed and distance. Local overalp would have:

- a distance of <= 30 km between stations
- a speed > 120 km/h

...and false positives: 

- any remaining detections with speeds > 120 km/h

False positives were removed

Local overlap detections were corrected by calculating the centroid between the two stations, replacing the coordinates with this new centroid location, recalculating the distance, replacing the name of the current station as a combination to the two stations, and for the next detection in sequence, replacing the previous station information with the new centroid information.

``` {r echo = FALSE}

# flag false positives 
final_cleaned_df <- cleaned_backwards %>%
  mutate(
    track_status = case_when(
      # flag as local overlapping detections with high speed and low distance
      travelDistanceKMs <= 30 & speedKmH > 120 ~ "LOCAL_OVERLAP",
      # flag as false positive for very high speeds 
      speedKmH > 120                           ~ "FALSE_POSITIVE",
      # all other flights flagged as valid
      TRUE                                     ~ "VALID_FLIGHT"
    )
  ) %>%
# arrange chronologically
  arrange(tagDeployID, tsStart_dt) %>%
  group_by(tagDeployID) %>%
  
  # Calculate midpoints for local overlaps
  mutate(
    lon = if_else(track_status == "LOCAL_OVERLAP" & !is.na(lon_previous), 
                  (lon + lon_previous) / 2, lon),
    lat = if_else(track_status == "LOCAL_OVERLAP" & !is.na(lat_previous), 
                  (lat + lat_previous) / 2, lat),
    stationName = if_else(
      track_status == "LOCAL_OVERLAP", 
      paste0(previousStationName, " & ", stationName), 
      stationName
    )
  ) %>%
  
  # arrange chronologically
  arrange(tagDeployID, tsStart_dt) %>%
  group_by(tagDeployID) %>%
  
  # calculate centroids using geosphere
  mutate(
    centroid = if_else(
      track_status == "LOCAL_OVERLAP" & !is.na(lon_previous),
      #centroid requires a matrix
      geosphere::midPoint(cbind(lon_previous, 
                                lat_previous), 
                          cbind(lon, lat)),
      cbind(lon, lat)
    ),
    lon = centroid[, 1],
    lat = centroid[, 2],
    
    stationName = if_else(
      track_status == "LOCAL_OVERLAP", 
      paste0(previousStationName, " & ", stationName), 
      stationName
    )
  ) %>%
  select(-centroid) %>% # remove temporary matrix column
  
  # Put centroid coordinates to the next sequential row
  mutate(
    follows_overlap = lag(track_status == "LOCAL_OVERLAP", 
                          default = FALSE),
    
    previousStationName = if_else(follows_overlap, 
                                  lag(stationName), 
                                  previousStationName),
    previousStationID   = if_else(follows_overlap, 
                                  lag(stationID), 
                                  previousStationID),
    lon_previous        = if_else(follows_overlap, 
                                  lag(lon), 
                                  lon_previous),
    lat_previous        = if_else(follows_overlap, 
                                  lag(lat), 
                                  lat_previous),
    
    # Update track_status 
    track_status = if_else(follows_overlap & 
                             track_status == "VALID_FLIGHT", 
                           "LOCAL_OVERLAP_RESOLVED", 
                           track_status)
  ) %>%
  
  # Remove the duplicate row, unless it is the only one
  filter(track_status != "LOCAL_OVERLAP" | lead(follows_overlap, 
                                                default = FALSE) == FALSE) %>%
  
  # Recalculate distances for the centroids
  mutate(
    distance_km = if_else(
      follows_overlap,
      geosphere::distHaversine(cbind(lon_previous, 
                                     lat_previous), 
                               cbind(lon, lat)) / 1000,
      distance_km
    )
  ) %>%
  select(-follows_overlap) %>% 
  ungroup()

table(final_cleaned_df$track_status)
```

Number of detections after removing false positives and resolving local overlap
```{r echo = FALSE}
nrow(final_cleaned_df)
```

## Flight information added

Flight specific information was added as additional columns
This includes:

- flight number (sequential flights per individual)
- speed per step in the flight
- turning angle (where it could be calculated)
- diel period (daylight or night) of the flight
- hours from sunrise and sunset (can be used to identify flights close to take-off or landing)
- additional grouping variable added:
  - which subbasin the occurrence took place in within the GLWS

## Individual information added

Information about each individual was added including:

- the tag deployment coordinates
- sex/age of the individual (if the project included this)
- days post migration the detection occurred
- tagging location potential issue flags
  - whether the tagging location was inside or outside the GLWS
  - how far from the boundary and the closest lake shoreline the tagging location was

## Data Flags

There are two columns of data flags 

1) Valid Distance:
- for station pairs that are < 30 km apart = FALSE_LESS
- for station pairs that are > 965 km apart = FALSE_MORE

Pairs that are close together are likely local overlaps
Pairs that are too far apart are likely false positives 

``` {r echo = FALSE}
animalFlags <- read.csv("StationPairsFiltered.csv")

table(animalFlags$valid_distance)
```


2) Tagging location flags for spring phenology
- if the tagging location is within the GLWS = Flag_1
- if the tagging location is within the GLWS and far inland >5 km = Flag_2
- if the tagging location is within the GLWS and not during migration (was tagged in summer or winter) = Flag_3 

These are flagged to be investigated as potential issues for assessing spring arrival phenology

```{r echo = FALSE}
animalFlags <- read.csv("StationPairsFiltered.csv")

table(animalFlags$tagSite_Flags)
```
