---
title: "Station Summary Data"
author: "Amanda Liczner"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(motus)
library(lubridate)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(sf)

Sys.setenv(TZ = "UTC") 

```

Overview of the data from Motus Project #1 summary data (downloaded off the web). 

### Data preparation summary
The summary data was downloaded from all Motus Project #1 towers.
The data are downloaded as "pairs" with the current and previous station noted

The downloaded data included the following headings:
```{r echo=FALSE}
data_active <- read.csv("./StationDownloads/Motus-data-project-1_detections_downloaded-2026-06-03.csv")
#at first just did active towers to prevent crashing
data_inactive <- read.csv ("./StationDownloads/Motus-data-multiple-stations_detections_downloaded-2026-06-09.csv")

#the inactive data had frequency as num, convert so it matches
data_inactive <- data_inactive %>%  
  mutate(tagFrequency = as.character(tagFrequency))

#join active and inactive stations together

data_raw <- bind_rows(data_active, data_inactive) %>%
  select(-tagFrequency) %>%  # some rows have it some do not
  distinct()
str(data_raw)
```
**Additional columns were added:**

- calculated my own movement_duration_hous, traveDurationHours is from the download
- calculated my own distance_km, travelDistanceKMs is from the download
- added in a column for season
  - spring = April 15 - June 15
  - summer = June 15 - August 18
  - fall = August 15 - November 15
  - Winter = November 15- April 15
- sunrise_utc/sunset_utc: the timing of sunrise/sunset in UTC
- sunrise_local/sunset_local: the timing of sunrise/Sunset in Eastern Time
- current_in_GLWS/previous_in_GLWS: if the station (current or previous) is within the Great Lakes Watershed
- distance_km: the distance between the two stations
- valid_distance: 
  - FALSE_LESS = the distance between stations is less than 30 km
  - FALSE_MORE = the distance between stations is more than 965 km 
  - TRUE = the distance between stations is between 30 - 965 km
- flight:
  flight = when the detection between stations is < 12 hrs
  incidence = when the detection between stations is > 12 hrs
- stationDuplicateFlag: flags detection issues (note buffer of 3 seconds was used to account for potential delays in detection and processing time at stations)
  - flag_1 = the start and end time are the same
  - flag_2 multiple towers detected at once, differing start times between rows going back in time
  - flag_3 mismatched run where it ends before its start time. 
  - flag_1_resolved = where flag_1 issues were corrected 
- eventClusterID = unique identifier assigned to a group of overlapping or continuous tracking events for a specific individual, created while tracking flagged errors
- new_flight = to identify the start of a new flight, used while tracking flagged errrors, a new flight is when the movement duration is > 12 hrs
- flight_number: a sequential numbering of flights for that individual
- flight_ID: adds "Flight_" to the flight number (from flight_number)
- step_speed_kmh: the speed in km/h at that flight step
- turning_angle: the turning angle calculated, where possible (=> detections needed)
- diel_period: whether the flight is occuring at night or during daylight
- nearSun: whether the flight occured within one hour of sunrrise/sunset
  - sunrise = the flight occurred within one hour of sunrise
  - sunset = the flight occurred within on hour of sunset
  - none = the flight did not occur within one hour of sunrise or sunset. 

**Data removed:** 

- Detections at the same current and previous station were removed as we are focusing on flights
- If the previous station name is unknown, the detection was removed

```{r echo=FALSE}
station_pairs <- read.csv("observationSummary.csv")
filtered_pairs <- station_pairs %>% 
  filter(
    previousStationID != stationID,
    previousStationName != "Unknown",
    previousStationName != "Unknown station"
  )
cat("The number of detections before removals:", 
    nrow(station_pairs), "\n")

cat("The number of detections after removing same station and unknown stations:", 
    nrow(filtered_pairs), "\n")
cat("The number of detections removed:", (nrow(station_pairs)- nrow(filtered_pairs)))
```
- Only kept station pairs where at least one station (either current or previous) was within the Great Lakes Watershed

``` {r echo=FALSE}
#loading great lakes watershed polygon, contains subbasins for each lake
GLWatershed <- st_read("./greatlakes_subbasins/greatlakes_subbasins.shp")

#transform both the polygon and station pairs to lambert conformal conic

#need to run current and previous stations separate;y
current_sf <- st_as_sf(filtered_pairs, 
                       coords = c("lon", "lat"), 
                       crs = 4326) %>%
  st_transform(crs = 3348)

previous_sf <- st_as_sf(filtered_pairs, 
                        coords = c("lon_previous", "lat_previous"), 
                        crs = 4326) %>%
  st_transform(crs = 3348)

GLWS_proj <- st_transform(GLWatershed, crs = 3348)

# check if either of the stations are within the GL watershed, 
filtered_pairs$current_in_GLWS  <- lengths(st_intersects(current_sf, 
                                                         GLWS_proj)) > 0
filtered_pairs$previous_in_GLWS <- lengths(st_intersects(previous_sf, 
                                                         GLWS_proj)) > 0

# calculate distance between points
filtered_pairs$distance_km <- as.numeric(st_distance(current_sf, 
                                                     previous_sf, 
                                                     by_element = TRUE)
) / 1000 #convert to km

filtered_pairs$valid_distance <- ifelse(filtered_pairs$distance_km < 30, 
                                        "FALSE_LESS",
                                        ifelse(filtered_pairs$distance_km > 965, 
                                               "FALSE_MORE", 
                                               "TRUE"))

filtered_df <- filtered_pairs %>%
  filter(
    # Keep if either the current OR the previous station is in GLWS
    (current_in_GLWS | previous_in_GLWS))
cat("The number of detections after removing those not in the Great Lakes Watershed:",
    nrow(filtered_df))
cat("The number of detections removed:", 
    (nrow(filtered_pairs)-nrow(filtered_df)))

```
- detections under Flag_2 that have mismatched end times, and all Flag_3s were removed

### Correcting station flag issues

**Flag 1, the same start and end time between two different towers**
There were 26 detections with Flag_1

Detections under this flag occurred when two stations (confirmed it was only ever 2 stations) picked up the same individual was in range of two stations. 
To fix this issue, the midpoint of the two stations was calculated. This location became the new coordinates for the current station. The current station name was replaced with a combination of both station names separated by "_". The previous station of the next detection was renamed with the combined name. These corrections were flagged as "flag_1_resolved"

**Flag 2, multiple towers detected the individual once, differing start times between rows going back in time**

There were 1774 detections with Flag_2

Detections under this flag occurred when stations that are close together were picking up the same individual simultaneously and this looks like it resulted in a processing error where the detection was recorded twice (identical on 2 rows) _except_ once row would have the end time as earlier than the start time. 

To fix this error, I removed the duplicate row with the incorrect end time

**Flag 3, mismatched run where it ends before its start time**

There were 5527 detections with Flag_3

Detections here are similar to Flag_2 but there isn't a duplicated row. These detections appear to be mistakes that occurred during processing where the end time is recorded as before the start time. 

I could not figure out how to fix this error so these detections were removed. 



```{r }
flightInfo_df <- read.csv("StationPairsFiltered.csv")
str(flightInfo_df)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
