---
title: "Station Summary Data"
author: "Amanda Liczner"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(motus)
library(lubridate)
library(dplyr)
library(purrr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(sf)
library(ggplot2)
library(knitr)
library(mapview)
library(DT)

Sys.setenv(TZ = "UTC") 

```

Overview of the data from Motus Project #1 summary data (downloaded off the web). 

### Data preparation summary
The summary data was downloaded from all Motus Project #1 towers.
The data are downloaded as "pairs" with the current and previous station noted

A preview downloaded data:
```{r echo=FALSE, warning=FALSE}
data_ON <- read.csv("./StationDownloads/Motus-data-region-CA-ON_detections_downloaded-2026-06-23.csv")
data_NY <- read.csv("./StationDownloads/Motus-data-region-US-NY_detections_downloaded-2026-06-23.csv")
data_PA <- read.csv("./StationDownloads/Motus-data-region-US-PA_detections_downloaded-2026-06-23.csv")
data_OH <- read.csv("./StationDownloads/Motus-data-region-US-OH_detections_downloaded-2026-06-23.csv")
data_IN <- read.csv("./StationDownloads/Motus-data-region-US-IN_detections_downloaded-2026-06-23.csv")
data_IL <- read.csv("./StationDownloads/Motus-data-region-US-IL_detections_downloaded-2026-06-23.csv")
data_WI <- read.csv("./StationDownloads/Motus-data-region-US-WI_detections_downloaded-2026-06-23.csv")

# join all the station data 
data_raw <- bind_rows(data_ON,
                      data_NY,
                      data_PA,
                      data_OH,
                      data_IN,
                      data_IL,
                      data_WI) %>%  # some rows have it some do not
  distinct() 

# Format the date columns using lubridate functions
data_cleaned <- data_raw %>%
  mutate(
    tsStart_dt = dmy_hms(tsStart, tz = "UTC"),
    tsEnd_dt   = dmy_hms(tsEnd, tz = "UTC"),
    year_start = year(tsStart_dt),
    year_end   = year(tsEnd_dt)
    )

#================================================================
# Station metadata json file
#================================================================

# Station metadata json file with GPS coordinates
stationsJSON <- jsonlite::read_json("./JSONFiles/Stations.json")

stationsJSON <- jsonlite::fromJSON("./JSONFiles/Stations.json", 
                                   simplifyDataFrame = TRUE)

readr::write_csv(stationsJSON$results, "StationsJSON.csv")

#getting coordinates from stationsJSON to make a lookup table for 
# previous tower coordinates
id_coords_lookup <- stationsJSON$results %>% 
  select(stationID, 
         latitude, 
         longitude) %>% 
  filter(!is.na(latitude) & !is.na(longitude)) %>% 
  # Convert stationID to integer
  mutate(stationID = as.integer(stationID)) %>% 
  # Group by ID and take the first coordinate 
  group_by(stationID) %>% 
  summarise(
    latitude = first(latitude),
    longitude = first(longitude),
    .groups = "drop"
  )

# Add previous station coordinates using the ID columns
summary_coords <- data_cleaned %>% 
  # Convert both ID columns to integers
  mutate(
    previousStationID = as.integer(previousStationID),
    stationID         = as.integer(stationID)
  ) %>% 
  
  # Join the lookup table to the previousStationID column
  left_join(
    id_coords_lookup, 
    by = c("previousStationID" = "stationID")
  ) %>% 
  
  # Rename the new columns coordinates
  rename(
    lat_previous = latitude,
    lon_previous = longitude
  )

head(summary_coords)
```

**Additional columns were added:**

- column for season
  - spring = April 15 - June 15
  - summer = June 15 - August 18
  - fall = August 15 - November 15
  - Winter = November 15- April 15

```{r echo=FALSE}

obs_sum <- read.csv("observationSummary.csv")

table(obs_sum$season)
```

- column for sunrise/sunset time
- column for sunrise/sunset local time
- column for movement duration in hours

**number of detections in raw data**

```{r echo=FALSE}
nrow(obs_sum)
```

## Data cleaning

Number of detections after removing unknown stations and station pairs that were the same
(bird remained at the same station)

```{r echo=FALSE}
station_pairs <- read.csv("observationSummary.csv") #142303 obs

#filter out same station pairs and station unknown

filtered_pairs <- station_pairs %>% 
  filter(
    previousStationID != stationID,
    previousStationName != "Unknown",
    previousStationName != "Unknown station"
  )
nrow(filtered_pairs)
```

Number of detectons with at least one station in the station-pair within the Great Lakes
Watershed (GLWS)
```{r echo=FALSE}
#loading great lakes watershed polygon, contains subbasins for each lake
GLWatershed <- st_read("./greatlakes_subbasins/greatlakes_subbasins.shp")
#checking CRS
st_crs(GLWatershed) #ESPG 6269

#transform both the polygon and station pairs to lambert conformal conic

#need to run current and previous stations separately
current_sf <- st_as_sf(filtered_pairs, 
                       coords = c("lon", "lat"), 
                       crs = 4326) %>%
  st_transform(crs = 3348)

previous_sf <- st_as_sf(filtered_pairs, 
                        coords = c("lon_previous", "lat_previous"), 
                        crs = 4326) %>%
  st_transform(crs = 3348)

GLWS_proj <- st_transform(GLWatershed, crs = 3348)

# check if either of the stations are within the GL watershed, 
filtered_pairs$current_in_GLWS  <- lengths(st_intersects(current_sf, 
                                                         GLWS_proj)) > 0
filtered_pairs$previous_in_GLWS <- lengths(st_intersects(previous_sf, 
                                                         GLWS_proj)) > 0

# calculate distance between points
filtered_pairs$distance_km <- as.numeric(st_distance(current_sf, 
                                                     previous_sf, 
                                                     by_element = TRUE)
) / 1000 #convert to km

# Mark TRUE if distance is 30 - 965 km, otherwise
#FALSE_LESS if distance is less than 30 km 
# FALSE_MORE if distance is more than 965 km

filtered_pairs$valid_distance <- ifelse(filtered_pairs$distance_km < 30, 
                                        "FALSE_LESS",
                                        ifelse(filtered_pairs$distance_km > 965, 
                                               "FALSE_MORE", 
                                               "TRUE"))

#==================================================================
# station pair filtering by criteria
#==================================================================

filtered_df <- filtered_pairs %>%
  filter(
    # Keep if either the current OR the previous station is in GLWS
    (current_in_GLWS | previous_in_GLWS))

nrow(filtered_df) #35865  obs
```

